---
title: "Coding Exercise: Best Predictor via One-Way ANOVA"
subtitle: "Instructions, Hints, and Full Solution"
author: "Applied Statistics"
date: "May 12, 2026"
output:
  html_document:
    toc: true
    toc_depth: 2
---

# Exercise Instructions

You are given a classification-style setup where the target is a categorical group variable.

Your goal is to identify which numeric feature has the strongest group-separation signal using **One-Way ANOVA**.

## Requirements

1. Create a synthetic dataframe with **200 observations**.
2. Include one categorical variable named `Treatment_Group` with **3 levels**.
3. Include four continuous numeric features with different mean shifts and/or variances:
   - `Feature_1`
   - `Feature_2`
   - `Feature_3`
   - `Feature_4`
4. Use an iterative approach (`for` loop or `lapply`) to run `aov()` for each feature against `Treatment_Group`.
5. Extract the **F-statistic** and **p-value** from each ANOVA model.
6. Build a comparison table and rank features by descending F-statistic.
7. Select the top-ranked feature as the one with the strongest predictive separation.
8. Create a `ggplot2` boxplot of the top feature by `Treatment_Group`.

## ANOVa theory

In ANOVA, you compare between-group variation to within-group variation:

$$
F = \frac{\text{variance between groups}}{\text{variance within groups}}
$$

A higher F-statistic indicates clearer group separation relative to noise.

---

# Hints

- Build formulas dynamically with:

```{r, eval=FALSE}
as.formula(paste(feature_name, "~ Treatment_Group"))
```

- Extract ANOVA metrics from `summary(aov_model)[[1]]`:

```{r, eval=FALSE}
fit <- aov(feature ~ Treatment_Group, data = df)
sm <- summary(fit)[[1]]
f_val <- sm["Treatment_Group", "F value"]
p_val <- sm["Treatment_Group", "Pr(>F)"]
```

- Rank by descending F-statistic:

```{r, eval=FALSE}
results <- results[order(-results$F_statistic), ]
```

- Use a dynamic y mapping in `ggplot2`:

```{r, eval=FALSE}
aes(y = .data[[top_feature]])
```

---

# Full Solution Code (Commented)

```{r setup, message=FALSE, warning=FALSE}
# Load required package
library(ggplot2)

# Ensure reproducibility
set.seed(123)
```
## 1) Generate synthetic data

```{r generate-data}

n <- 200

# Grouping variable with 3 levels
Treatment_Group <- factor(
  sample(c("Control", "Treatment_A", "Treatment_B"), n, replace = TRUE)
)

# Group-specific means for each feature
# Feature_1: strongest separation
# Feature_2: moderate separation
# Feature_3: weak separation with high noise
# Feature_4: very weak separation
mu_f1 <- c(Control = 50, Treatment_A = 62, Treatment_B = 74)
mu_f2 <- c(Control = 30, Treatment_A = 35, Treatment_B = 40)
mu_f3 <- c(Control = 100, Treatment_A = 102, Treatment_B = 104)
mu_f4 <- c(Control = 10, Treatment_A = 10.5, Treatment_B = 11)

# Build numeric features with varying within-group variance
Feature_1 <- rnorm(n, mean = mu_f1[Treatment_Group], sd = 4)
Feature_2 <- rnorm(n, mean = mu_f2[Treatment_Group], sd = 8)
Feature_3 <- rnorm(n, mean = mu_f3[Treatment_Group], sd = 14)
Feature_4 <- rnorm(n, mean = mu_f4[Treatment_Group], sd = 6)

# Final dataframe
df <- data.frame(
  Treatment_Group,
  Feature_1,
  Feature_2,
  Feature_3,
  Feature_4
)

# Preview
head(df)
```

## 2) Iteratively run ANOVA for each numeric feature

```{r iterative-anova}

feature_names <- c("Feature_1", "Feature_2", "Feature_3", "Feature_4")

anova_list <- lapply(feature_names, function(feat) {
  # Dynamic formula: feat ~ Treatment_Group
  fml <- as.formula(paste(feat, "~ Treatment_Group"))
  fit <- aov(fml, data = df)
  sm <- summary(fit)[[1]]

  data.frame(
    Feature = feat,
    F_statistic = sm["Treatment_Group", "F value"],
    p_value = sm["Treatment_Group", "Pr(>F)"],
    row.names = NULL
  )
})

anova_results <- do.call(rbind, anova_list)
anova_results
```
## 3) Rank features by F-statistic (highest = strongest signal)

```{r rank-features}

anova_results <- anova_results[order(-anova_results$F_statistic), ]
anova_results$Rank <- seq_len(nrow(anova_results))

anova_results <- anova_results[, c("Rank", "Feature", "F_statistic", "p_value")]
anova_results

# Top-performing feature
top_feature <- anova_results$Feature[1]
cat("Top feature by ANOVA F-statistic:", top_feature, "\n")
```
## 4) Visualize top-performing feature
```{r plot-top-feature, fig.width=8, fig.height=4.5}

ggplot(df, aes(x = Treatment_Group, y = .data[[top_feature]], fill = Treatment_Group)) +
  geom_boxplot(alpha = 0.8) +
  labs(
    title = paste("Top Feature by One-Way ANOVA:", top_feature),
    x = "Treatment Group",
    y = top_feature
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

---

1. Why can a feature with larger mean differences still fail to rank first?
2. How does within-group variance affect the F-statistic?
3. Would scaling a feature (e.g., multiplying by 10) change ANOVA conclusions?
