---
title: "Applied Logistic Regression Workshop"
author: "Applied Statistics"
date: "2026-05-06"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme: readable
---

# Workshop Notebook: Applied Logistic Regression

**Dataset:** `bank_loan_data.csv`  
**Business problem:** Predict loan default using credit card balance and student status.  
**Goal:** Build, interpret, and evaluate a logistic regression model for loan default.

## Data Generator

This first chunk recreates the dataset used in the workshop. It generates a known relationship between balance, student status, and default so the fitted model can be checked against the lecture values.

```{r}
set.seed(2026)

N <- 5000

Student <- rbinom(N, size = 1, prob = 0.3)

Balance <- round(rnorm(N, mean = 1200, sd = 500))
Balance[Balance < 0] <- 0
Balance[Balance > 3500] <- 3500

true_log_odds <- -8.5 + (0.0054 * Balance) - (0.60 * Student)

true_probability <- 1 / (1 + exp(-true_log_odds))

Default <- rbinom(N, size = 1, prob = true_probability)

df_bank <- data.frame(
  Default = Default,
  Balance = Balance,
  Student = Student
)

write.csv(df_bank, "bank_loan_data.csv", row.names = FALSE)
cat("'bank_loan_data.csv' has been generated in your working directory!\n\n")

cat("bank_loan_data.csv has been written to the working directory.\n\n")

cat("--- BASELINE STATS ---\n")
cat("Total Customers:", nrow(df_bank), "\n")
cat("Overall Default Rate:", round(mean(df_bank$Default) * 100, 1), "%\n\n")

cat("--- MODEL VERIFICATION ---\n")
test_model <- glm(Default ~ Balance + Student, data = df_bank, family = "binomial")

print(round(coef(test_model), 4))

bank_loan_data <- df_bank
```

## Section 0: Environment Setup & Data Ingestion

This section loads the packages, reads the dataset, and checks the class balance before any model is fitted. The default rate tells us how imbalanced the target is and gives a baseline for later comparison.

```{r}
library(dplyr)
library(ggplot2)
library(pROC)

data_path <- "bank_loan_data.csv"
if (!file.exists(data_path)) {
  stop("Could not find bank_loan_data.csv. Place it in the same folder as this workshop notebook.")
}

bank_loan_data <- read.csv(data_path, stringsAsFactors = FALSE)
```

### Task

Run `head()` and `summary()` on the dataset to explore the variables. Then check the baseline default rate:

```{r}
head(bank_loan_data)
summary(bank_loan_data)
mean(bank_loan_data$Default == 1, na.rm = TRUE)
```

## Section 1: The OLS Catastrophe (Proving the Theory)

Ordinary Least Squares produces values on the real line, which makes sense for a continuous outcome but not for a probability. This section shows the failure directly with a binary target.

### Task

Fit a standard linear model using `lm(Default ~ Balance)`.

Create a test dataframe with two extreme customers:
- Customer A: `Balance = 3500`
- Customer B: `Balance = 0`

Use `predict()` to estimate their default risk.

```{r}
ols_model <- lm(Default ~ Balance, data = bank_loan_data)
summary(ols_model)

test_customers <- data.frame(
  Balance = c(3500, 0)
)

ols_predictions <- predict(ols_model, newdata = test_customers)
ols_predictions
```


Look at the predictions for Customer A and Customer B. Why are these numbers completely invalid for a business report?

## Section 2: The Logistic Engine

Logistic regression keeps predictions inside the probability range by modeling the log-odds and estimating the coefficients with maximum likelihood.

### Task

Fit a logistic regression model using `glm(..., family = binomial(link = "logit"))`.
Print the `summary()` of the model.

```{r}
logit_model <- glm(Default ~ Balance, data = bank_loan_data, family = binomial(link = "logit"))
summary(logit_model)
```


Look at the coefficient for `Balance` (for example, `0.0054`). Does this mean that every additional dollar adds 0.54% to the probability of default? If not, what units is this coefficient actually in?

## Section 3: Translation & The Golden Sentence

The coefficient lives on the log-odds scale, which is hard to explain in business terms. Converting it with `exp()` turns it into an odds ratio.

### Task

Extract the coefficient for `Balance`.
Use the `exp()` function to exponentiate the coefficient.

```{r}
balance_coef <- coef(logit_model)["Balance"]
balance_odds_ratio <- exp(balance_coef)

balance_coef
balance_odds_ratio
```


Write the "Golden Sentence" for this finding. For example: "For every $1 increase in balance, the odds of default are ..."

## Section 4: Multiple Predictors (Controlling for Groups)

Balance is not the only driver of default risk. Student status adds a second effect and lets the model compare borrowers at the same balance.

### Task

Fit a new model: `glm(Default ~ Balance + Student)`.
Exponentiate the `Student` coefficient.

```{r}
if (!"Student" %in% names(bank_loan_data)) {
  stop("This workshop expects a Student column in bank_loan_data.csv.")
}

logit_model_student <- glm(Default ~ Balance + Student, data = bank_loan_data, family = binomial(link = "logit"))
summary(logit_model_student)

student_coef <- coef(logit_model_student)["Student"]
student_odds_ratio <- exp(student_coef)

student_coef
student_odds_ratio
```

The exponentiated coefficient for `Student` is roughly `0.54`. Because this number is less than 1, what does it mean for a student's risk compared to a non-student, assuming they have the exact same credit card balance?

## Section 5: Generating Probabilities & Business Decisions

A fitted model becomes useful when it produces probabilities that can be converted into approve or deny decisions. This section turns model output into a policy rule.

### Task

Use `predict(model, type = "response")` to generate probabilities for the entire dataset.  
Instructor note: remind yourself what happens if you forget `type = "response"`.

Create a new column called `Decision_50`. Use `ifelse()` to predict 1 (Deny) if the probability is greater than 0.50, and 0 (Approve) if it is less than 0.50.
Generate a confusion matrix using `table()`.

```{r}
bank_loan_data$Predicted_Probability <- predict(logit_model_student, newdata = bank_loan_data, type = "response")

bank_loan_data$Decision_50 <- ifelse(bank_loan_data$Predicted_Probability > 0.50, 1, 0)

conf_matrix_50 <- table(
  Predicted = bank_loan_data$Decision_50,
  Actual = bank_loan_data$Default
)
conf_matrix_50
```

Identify the False Positives and False Negatives in your matrix. If a defaulted loan costs the bank $10,000, and a lost good customer only costs the bank $500 in lost interest, is `0.50` a smart threshold?

## Section 6: Threshold Optimization

The threshold determines how the bank balances missed defaults against rejected good borrowers. Lowering the cutoff changes that trade-off.

### Task

Create a new column called `Decision_15` using a threshold of `0.15`.
Generate a new confusion matrix.

```{r}
bank_loan_data$Decision_15 <- ifelse(bank_loan_data$Predicted_Probability > 0.15, 1, 0)

conf_matrix_15 <- table(
  Predicted = bank_loan_data$Decision_15,
  Actual = bank_loan_data$Default
)
conf_matrix_15
```

Compare this matrix to the previous one. What happened to our False Negatives (defaulters we missed)? What happened to our False Positives (good customers we angered)? Explain the trade-off.

## Section 7: The Ultimate Grade (ROC & AUC)

ROC and AUC summarize model quality across every possible threshold instead of only one cutoff. They show how well the scores separate defaulters from non-defaulters.

### Task

Use the `roc()` function from the `pROC` package.
Use `plot()` to visualize the ROC curve.
Use `auc()` to calculate the Area Under the Curve.

```{r}
roc_obj <- roc(response = bank_loan_data$Default, predictor = bank_loan_data$Predicted_Probability)
plot(roc_obj, main = "ROC Curve for Logistic Regression", col = "steelblue", lwd = 2)
auc_value <- auc(roc_obj)
auc_value
```

What is the model's AUC score? Based on the `0.5` (Coin Toss) to `1.0` (Psychic) scale, how would you rate the quality of this classifier?

## Section 8: The "Boss's Desk" Challenge (Independent Practice)

This final exercise asks for a decision on a new applicant using the fitted model. The result should be a probability first and a business recommendation second.

**Scenario:** A new applicant just walked into the branch. They are NOT a student, and they have a credit card balance of $2,200.

**Instructions:** Write the code to predict their exact probability of default. Based on a conservative 20% risk threshold, do you approve or deny their loan?

**Output expected:** Create the new data frame, run the prediction, and write a one-sentence business recommendation.

```{r}
new_applicant <- data.frame(
  Balance = 2200,
  Student = 0
)

new_applicant_probability <- predict(logit_model_student, newdata = new_applicant, type = "response")
new_applicant_probability

if (new_applicant_probability > 0.20) {
  recommendation <- "Deny the loan: the predicted default risk is above the 20% threshold."
} else {
  recommendation <- "Approve the loan: the predicted default risk is below the 20% threshold."
}

recommendation
```

### Final Reflection

Write one sentence for your manager: What is the predicted probability, and should the loan be approved or denied?
