---
title: "Applied Statistics Lesson 7: Clustering & Distance Metrics"
author: "Lazarski University"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)
```

This setup chunk configures `knitr` options and plotting defaults used throughout the notebook (figure size, DPI, and message/warning behaviour).

# 1. Setup & Data Preparation

## Load Required Libraries
# Load the libraries used for plotting, clustering, and distance calculations.
```{r}
# Core tidyverse
library(ggplot2)
library(dplyr)
library(patchwork)
library(tidyr)

# Clustering
library(cluster)      # silhouette, daisy
library(ggdendro)     # dendrogram visualization

# Distance metrics
library(proxy)        # distance calculations
library(clusterSim)   # cluster validity indices (index.DB, index.G1)

set.seed(42)
```

## Generate Synthetic Dataset

We'll create the same 120-point, 3-cluster dataset used in the lesson slides.

```{r}
# Generate synthetic 3-cluster data
n_per_cluster <- 40
clusters_3 <- data.frame(
  x = c(
    rnorm(n_per_cluster, mean = 2, sd = 0.6),
    rnorm(n_per_cluster, mean = 7, sd = 0.6),
    rnorm(n_per_cluster, mean = 7, sd = 0.6)
  ),
  y = c(
    rnorm(n_per_cluster, mean = 2, sd = 0.6),
    rnorm(n_per_cluster, mean = 7, sd = 0.6),
    rnorm(n_per_cluster, mean = 2, sd = 0.6)
  ),
  true_cluster = rep(1:3, each = n_per_cluster)
)

# Display data summary
head(clusters_3, 10)
summary(clusters_3[, 1:2])

# Visualize true clusters
ggplot(clusters_3, aes(x = x, y = y, color = factor(true_cluster))) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = c("#1A3A6B", "#4A90D9", "#B03A2E")) +
  labs(
    title = "True Clustering (3 Clusters)",
    color = "True Cluster",
    x = "Feature 1 (x)",
    y = "Feature 2 (y)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

  The plot above shows the synthetic dataset with the true cluster labels; we use this dataset to demonstrate clustering algorithms and evaluation metrics.

---

# 2. Distance Metrics

## 2.1 Distance Matrix Calculation

### Euclidean Distance

**Formula:** $d_{Euclidean}(i,j) = \sqrt{\sum_{k=1}^{p} (x_{ik} - x_{jk})^2}$

```{r}
# Calculate Euclidean distance matrix
euclidean_dist <- dist(clusters_3[, 1:2], method = "euclidean")

# Display a subset
as.matrix(euclidean_dist)[1:5, 1:5]
```

This computes pairwise Euclidean distances between samples; the resulting matrix is symmetric and will be used by hierarchical clustering and silhouette calculations.

### Manhattan Distance

**Formula:** $d_{Manhattan}(i,j) = \sum_{k=1}^{p} |x_{ik} - x_{jk}|$

```{r}
# Calculate Manhattan distance matrix
manhattan_dist <- dist(clusters_3[, 1:2], method = "manhattan")

# Display a subset
as.matrix(manhattan_dist)[1:5, 1:5]
```

Manhattan (L1) distances are shown here for comparison; they measure absolute coordinate differences and may behave differently than Euclidean distances on the same data.

### Cosine Distance

**Formula:** $d_{Cosine}(i,j) = 1 - \frac{\mathbf{x}_i \cdot \mathbf{x}_j}{||\mathbf{x}_i|| \cdot ||\mathbf{x}_j||}$

```{r}
# Calculate Cosine distance using proxy package
cosine_dist <- dist(clusters_3[, 1:2], method = "cosine")

# Display a subset
as.matrix(cosine_dist)[1:5, 1:5]
```

Cosine distances focus on angular differences between vectors (orientation), which can be helpful when magnitude differences are not meaningful.

## 2.2 Distance Matrix Visualization

The following chunk prepares the Euclidean distance matrix for visualization and plots a heatmap to inspect pairwise similarities.

```{r}
# Convert to matrix for heatmap
dist_matrix <- as.matrix(euclidean_dist)

# Create heatmap using ggplot2
dist_df <- data.frame(
  i = rep(1:nrow(dist_matrix), times = ncol(dist_matrix)),
  j = rep(1:ncol(dist_matrix), each = nrow(dist_matrix)),
  distance = c(dist_matrix)
)

ggplot(dist_df, aes(x = j, y = i, fill = distance)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#0D1B3E") +
  labs(
    title = "Euclidean Distance Matrix Heatmap",
    x = "Sample Index",
    y = "Sample Index",
    fill = "Distance"
  ) +
  theme_minimal() +
  coord_fixed()
```

  The heatmap visually highlights which sample pairs are close or far in Euclidean space; block structure often indicates cluster membership.

## 2.3 Distance Comparison
# Compare numeric values from different distance metrics for a small set of pairs and compute correlations across methods.

```{r}
# Compare distances for the first 10 point pairs
comparison <- data.frame(
  Pair = 1:10,
  Euclidean = as.vector(euclidean_dist)[1:10],
  Manhattan = as.vector(manhattan_dist)[1:10],
  Cosine = as.vector(cosine_dist)[1:10]
)

print(comparison)

# Correlation between distance methods
cat("\nCorrelation between distance methods:\n")
dist_comparison <- cbind(
  Euclidean = as.vector(euclidean_dist),
  Manhattan = as.vector(manhattan_dist),
  Cosine = as.vector(cosine_dist)
)
cor(dist_comparison)
```

Comparing distances and their correlations helps understand how metric choice affects perceived similarity between points.

---

# 3. K-Means Clustering

## 3.1 Manual K-Means Implementation

Let's implement K-Means from scratch to understand the algorithm:

```{r}
# Manual K-Means function (robust to data.frames / matrices)
kmeans_manual <- function(data, k, max_iter = 100, tol = 1e-4, nstart = 1) {
  # Ensure numeric matrix input
  if (is.data.frame(data)) data <- as.matrix(data)
  if (!is.matrix(data)) stop("`data` must be a matrix or data.frame")
  if (!is.numeric(data)) stop("`data` must contain numeric values only")

  results <- list()
  best_wss <- Inf

  n_samples <- nrow(data)
  n_features <- ncol(data)

  for (start in 1:nstart) {
    # Initialize: random starting centroids (rows from data)
    centroid_indices <- sample(1:n_samples, k, replace = FALSE)
    centroids <- data[centroid_indices, , drop = FALSE]

    wss_history <- numeric(max_iter)

    for (iter in 1:max_iter) {
      # Step 1: Assign points to nearest centroid (squared Euclidean distances)
      distances <- sapply(1:k, function(j) rowSums((data - matrix(centroids[j, ], n_samples, n_features, byrow = TRUE))^2))
      assignments <- apply(distances, 1, which.min)

      # Step 2: Calculate within-cluster sum of squares (WSS)
      wss <- sum(sapply(1:k, function(c_idx) {
        cluster_points <- data[assignments == c_idx, , drop = FALSE]
        if (nrow(cluster_points) == 0) return(0)
        sum(rowSums((cluster_points - matrix(centroids[c_idx, ], nrow(cluster_points), n_features, byrow = TRUE))^2))
      }))
      wss_history[iter] <- wss

      # Step 3: Update centroids
      new_centroids <- t(sapply(1:k, function(c_idx) {
        cluster_points <- data[assignments == c_idx, , drop = FALSE]
        if (nrow(cluster_points) == 0) {
          centroids[c_idx, ]
        } else {
          colMeans(cluster_points)
        }
      }))

      # Check convergence (centroid movement)
      centroid_shift <- sum((new_centroids - centroids)^2)
      centroids <- new_centroids

      if (centroid_shift < tol) {
        wss_history <- wss_history[1:iter]
        break
      }
    }

    # Store results from this run if better
    if (wss < best_wss) {
      best_wss <- wss
      results$centroids <- centroids
      results$assignments <- assignments
      results$wss_history <- wss_history
      results$wss <- wss
    }
  }

  return(results)
}

# Run manual K-Means with k=3
km_manual <- kmeans_manual(clusters_3[, 1:2], k = 3, nstart = 10)

# View results
cat("Manual K-Means (k=3):\n")
cat("Final WSS:", km_manual$wss, "\n")
cat("Assignments:", table(km_manual$assignments), "\n")
```

## 3.2 Built-in K-Means Function

```{r}
# Use R's built-in kmeans function
km_builtin <- kmeans(clusters_3[, 1:2], centers = 3, nstart = 50, iter.max = 300)

# Summary
cat("Built-in K-Means (k=3):\n")
cat("Within-cluster sum of squares (WSS):", km_builtin$tot.withinss, "\n")
cat("Cluster sizes:", km_builtin$size, "\n")
print(km_builtin$centers)
```

## 3.3 K-Means Visualization: Step-by-Step

```{r}
# Create 2x2 visualization showing K-Means iterations
plot_kmeans_iteration <- function(data, km_result, iteration_title) {
  
  plot_data <- data.frame(
    x = data$x,
    y = data$y,
    cluster = factor(km_result$cluster),
    true_cluster = factor(data$true_cluster)
  )
  
  p <- ggplot(plot_data, aes(x = x, y = y, color = cluster)) +
    geom_point(size = 3, alpha = 0.7) +
    geom_point(
      data = data.frame(
        x = km_result$centers[, 1],
        y = km_result$centers[, 2],
        cluster = factor(1:nrow(km_result$centers))
      ),
      aes(fill = cluster),
      shape = 23,
      size = 8,
      color = "black",
      stroke = 2
    ) +
    scale_color_manual(values = c("#1A3A6B", "#4A90D9", "#B03A2E")) +
    scale_fill_manual(values = c("#1A3A6B", "#4A90D9", "#B03A2E")) +
    labs(
      title = iteration_title,
      x = "Feature 1",
      y = "Feature 2",
      color = "Cluster",
      fill = "Centroid"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  return(p)
}

# Visualize final K-Means result
plot_kmeans_iteration(clusters_3, km_builtin, "K-Means Clustering (k=3)")
```

## 3.4 Convergence Criteria Analysis

```{r}
# Explore convergence with different parameters
convergence_test <- data.frame(
  k = numeric(),
  converged_iter = numeric(),
  total_wss = numeric(),
  tol = numeric()
)

for (k_val in 2:6) {
  for (tol_val in c(1e-3, 1e-4, 1e-5)) {
    km_temp <- kmeans_manual(clusters_3[, 1:2], k = k_val, max_iter = 500, tol = tol_val, nstart = 5)
    convergence_test <- rbind(convergence_test, data.frame(
      k = k_val,
      converged_iter = length(km_temp$wss_history),
      total_wss = km_temp$wss,
      tol = tol_val
    ))
  }
}

print(convergence_test)

# Visualization: WSS vs iterations for different k values
wss_progress <- data.frame()
for (k_val in 2:5) {
  km_temp <- kmeans_manual(clusters_3[, 1:2], k = k_val, max_iter = 300, nstart = 10)
  wss_progress <- rbind(wss_progress, data.frame(
    iteration = 1:length(km_temp$wss_history),
    wss = km_temp$wss_history,
    k = k_val
  ))
}

ggplot(wss_progress, aes(x = iteration, y = wss, color = factor(k))) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~k, scales = "free_y") +
  scale_color_manual(values = c("#1A3A6B", "#4A90D9", "#B03A2E", "#C9A84C")) +
  labs(
    title = "K-Means: WSS Progress by Iteration",
    x = "Iteration",
    y = "Within-Cluster SS",
    color = "k"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

# 4. Hierarchical Clustering

## 4.1 Hierarchical Clustering with Different Linkages

```{r}
# Perform hierarchical clustering with different linkage methods
linkage_methods <- c("single", "complete", "average", "ward.D2")

hclust_results <- lapply(linkage_methods, function(method) {
  hclust(euclidean_dist, method = method)
})

names(hclust_results) <- linkage_methods

# Print dendrogram for Ward linkage
print(hclust_results$ward.D2)
```

## 4.2 Dendrogram Visualization

```{r}
# Single dendrogram for Ward linkage
ggdendrogram(hclust_results$ward.D2, theme_dendro = FALSE) +
  labs(
    title = "Hierarchical Clustering Dendrogram (Ward Linkage)",
    y = "Height (Distance)"
  ) +
  theme_minimal()
```

## 4.3 Linkage Methods Comparison

```{r}
# Create dendrograms for all linkage methods side-by-side
plots <- lapply(linkage_methods, function(method) {
  ggdendrogram(hclust_results[[method]], theme_dendro = FALSE) +
    labs(title = paste(toupper(method), "Linkage")) +
    theme_minimal() +
    theme(plot.title = element_text(size = 10, face = "bold"))
})

wrap_plots(plots, ncol = 2) +
  plot_annotation(title = "Comparison of Linkage Methods")
```

## 4.4 Cluster Assignment from Dendrogram

```{r}
# Cut dendrogram to get cluster assignments (k=3)
clusters_from_hclust <- cutree(hclust_results$ward.D2, k = 3)

# Visualize clusters from hierarchical clustering
hclust_data <- data.frame(
  x = clusters_3$x,
  y = clusters_3$y,
  cluster = factor(clusters_from_hclust)
)

ggplot(hclust_data, aes(x = x, y = y, color = cluster)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = c("#1A3A6B", "#4A90D9", "#B03A2E")) +
  labs(
    title = "Hierarchical Clustering Result (k=3, Ward Linkage)",
    x = "Feature 1",
    y = "Feature 2",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 4.5 Dendrogram with Cut Line

```{r}
# Dendrogram with horizontal cut line for k=3
dend_data <- dendro_data(hclust_results$ward.D2)

# Cut height for k=3
cut_height <- sort(hclust_results$ward.D2$height, decreasing = TRUE)[2]

ggplot() +
  geom_segment(data = segment(dend_data), aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_point(data = label(dend_data), aes(x = x, y = y), size = 3, alpha = 0.7) +
  geom_hline(yintercept = cut_height, linetype = "dashed", color = "red", size = 1.2) +
  labs(
    title = "Dendrogram with Cut Line (k=3)",
    y = "Height (Distance)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

---

# 5. Cluster Evaluation Metrics

## 5.1 Within-Cluster Sum of Squares (WSS)

**Formula:** $WSS = \sum_{i=1}^{k} \sum_{j \in C_i} ||x_j - \mu_i||^2$

```{r}
# Calculate WSS for k=2 to 10
wss_values <- numeric(10)
for (k in 1:10) {
  km_k <- kmeans(clusters_3[, 1:2], centers = k, nstart = 50)
  wss_values[k] <- km_k$tot.withinss
}

wss_df <- data.frame(
  k = 1:10,
  wss = wss_values
)

# Visualize WSS vs k
ggplot(wss_df, aes(x = k, y = wss)) +
  geom_line(size = 1, color = "#1A3A6B") +
  geom_point(size = 3, color = "#1A3A6B") +
  geom_vline(xintercept = 3, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = 3.2, y = max(wss_df$wss) * 0.95, label = "Elbow", size = 4, color = "red") +
  labs(
    title = "Elbow Method: Within-Cluster Sum of Squares",
    x = "Number of Clusters (k)",
    y = "Within-Cluster Sum of Squares (WSS)"
  ) +
  theme_minimal()

print(wss_df)
```

## 5.2 Silhouette Score: Calculation

**Formula for each point i:**
- $a(i)$ = average distance to points in same cluster
- $b(i)$ = minimum average distance to points in other clusters
- $s(i) = \frac{b(i) - a(i)}{max(a(i), b(i))}$

```{r}
# Calculate silhouette scores for different k values
silhouette_scores <- data.frame()

for (k in 2:6) {
  km_k <- kmeans(clusters_3[, 1:2], centers = k, nstart = 50)
  sil <- silhouette(km_k$cluster, euclidean_dist)
  
  silhouette_scores <- rbind(silhouette_scores, data.frame(
    k = k,
    avg_silhouette = mean(sil[, 3]),
    min_silhouette = min(sil[, 3]),
    max_silhouette = max(sil[, 3])
  ))
}

print(silhouette_scores)

# Visualize average silhouette score vs k
ggplot(silhouette_scores, aes(x = k, y = avg_silhouette)) +
  geom_line(size = 1.2, color = "#4A90D9") +
  geom_point(size = 4, color = "#4A90D9") +
  geom_vline(xintercept = which.max(silhouette_scores$avg_silhouette) + 1, 
             linetype = "dashed", color = "green", size = 1) +
  labs(
    title = "Silhouette Score vs k",
    x = "Number of Clusters (k)",
    y = "Average Silhouette Score"
  ) +
  ylim(0, 1) +
  theme_minimal()
```

## 5.3 Detailed Silhouette Analysis

```{r}
# Create silhouette plot for k=3
km_3 <- kmeans(clusters_3[, 1:2], centers = 3, nstart = 50)
sil_3 <- silhouette(km_3$cluster, euclidean_dist)

# Extract silhouette data
sil_df <- data.frame(
  index = 1:nrow(sil_3),
  silhouette = sil_3[, 3],
  cluster = sil_3[, 1]
)

# Sort by cluster and silhouette
sil_df <- sil_df %>% arrange(cluster, silhouette)
sil_df$index_ordered <- 1:nrow(sil_df)

# Plot silhouette widths
ggplot(sil_df, aes(x = index_ordered, y = silhouette, fill = factor(cluster))) +
  geom_col() +
  geom_hline(yintercept = mean(sil_3[, 3]), linetype = "dashed", color = "red", size = 1) +
  scale_fill_manual(values = c("#1A3A6B", "#4A90D9", "#B03A2E")) +
  labs(
    title = "Silhouette Plot for k=3 Clustering",
    x = "Points (sorted by cluster)",
    y = "Silhouette Width",
    fill = "Cluster"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank())

# Point-level silhouette interpretation
cat("\nSilhouette Score Interpretation for k=3:\n")
cat("Average Silhouette Score:", round(mean(sil_3[, 3]), 4), "\n")
cat("Points with s(i) > 0.5:", sum(sil_3[, 3] > 0.5), "\n")
cat("Points with s(i) < 0:", sum(sil_3[, 3] < 0), "\n")
```

## 5.4 Other Evaluation Metrics

### Davies-Bouldin Index

**Lower is better** (measures average similarity between each cluster and its most similar cluster)

```{r}
# Calculate Davies-Bouldin Index for k=2 to 6
davies_bouldin <- numeric(5)
for (i in 2:6) {
  km_i <- kmeans(clusters_3[, 1:2], centers = i, nstart = 50)
  davies_bouldin[i-1] <- index.DB(clusters_3[, 1:2], km_i$cluster)$DB
}

db_df <- data.frame(
  k = 2:6,
  davies_bouldin = davies_bouldin
)

print(db_df)

ggplot(db_df, aes(x = k, y = davies_bouldin)) +
  geom_line(size = 1.2, color = "#B03A2E") +
  geom_point(size = 4, color = "#B03A2E") +
  labs(
    title = "Davies-Bouldin Index vs k (Lower is Better)",
    x = "Number of Clusters (k)",
    y = "Davies-Bouldin Index"
  ) +
  theme_minimal()
```

### Calinski-Harabasz Index

**Higher is better** (ratio of between-cluster to within-cluster variance)

```{r}
# Calculate Calinski-Harabasz Index for k=2 to 6
calinski_harabasz <- numeric(5)
for (i in 2:6) {
  km_i <- kmeans(clusters_3[, 1:2], centers = i, nstart = 50)
  calinski_harabasz[i-1] <- index.G1(clusters_3[, 1:2], km_i$cluster)
}

ch_df <- data.frame(
  k = 2:6,
  calinski_harabasz = calinski_harabasz
)

print(ch_df)

ggplot(ch_df, aes(x = k, y = calinski_harabasz)) +
  geom_line(size = 1.2, color = "#C9A84C") +
  geom_point(size = 4, color = "#C9A84C") +
  labs(
    title = "Calinski-Harabasz Index vs k (Higher is Better)",
    x = "Number of Clusters (k)",
    y = "Calinski-Harabasz Index"
  ) +
  theme_minimal()
```

## 5.5 Comprehensive Metrics Comparison

```{r}
# Create comprehensive metrics table for k=2 to 5
metrics_table <- data.frame(
  k = 2:5,
  WSS = numeric(4),
  Silhouette = numeric(4),
  Davies_Bouldin = numeric(4),
  Calinski_Harabasz = numeric(4)
)

for (i in 1:4) {
  k_val <- i + 1
  km_temp <- kmeans(clusters_3[, 1:2], centers = k_val, nstart = 50)
  sil_temp <- silhouette(km_temp$cluster, euclidean_dist)
  
  metrics_table$WSS[i] <- round(km_temp$tot.withinss, 2)
  metrics_table$Silhouette[i] <- round(mean(sil_temp[, 3]), 4)
  metrics_table$Davies_Bouldin[i] <- round(index.DB(clusters_3[, 1:2], km_temp$cluster)$DB, 4)
  metrics_table$Calinski_Harabasz[i] <- round(index.G1(clusters_3[, 1:2], km_temp$cluster), 2)
}

print(metrics_table)

# Create visualization: compare all metrics (normalized)
metrics_normalized <- metrics_table
metrics_normalized$WSS <- 1 - (metrics_normalized$WSS / max(metrics_normalized$WSS))  # Invert
metrics_normalized$Davies_Bouldin <- 1 - (metrics_normalized$Davies_Bouldin / max(metrics_normalized$Davies_Bouldin))  # Invert

metrics_long <- metrics_normalized %>%
  pivot_longer(-k, names_to = "metric", values_to = "normalized_value")

ggplot(metrics_long, aes(x = k, y = normalized_value, color = metric, group = metric)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("#1A3A6B", "#4A90D9", "#B03A2E", "#C9A84C")) +
  labs(
    title = "Normalized Metrics Comparison (higher is better for all)",
    x = "Number of Clusters (k)",
    y = "Normalized Score",
    color = "Metric"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

# 6. K-Means vs Hierarchical Clustering

## Comparison

```{r}
# Cluster assignments from both methods (k=3)
km_3_clusters <- km_3$cluster
hclust_3_clusters <- cutree(hclust_results$ward.D2, k = 3)

# Create comparison dataframe
comparison_df <- data.frame(
  x = clusters_3$x,
  y = clusters_3$y,
  true_cluster = factor(clusters_3$true_cluster),
  kmeans_cluster = factor(km_3_clusters),
  hclust_cluster = factor(hclust_3_clusters)
)

# Side-by-side visualization
p1 <- ggplot(comparison_df, aes(x = x, y = y, color = kmeans_cluster)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = c("#1A3A6B", "#4A90D9", "#B03A2E")) +
  labs(
    title = "K-Means Clustering",
    x = "Feature 1",
    y = "Feature 2",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

p2 <- ggplot(comparison_df, aes(x = x, y = y, color = hclust_cluster)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = c("#1A3A6B", "#4A90D9", "#B03A2E")) +
  labs(
    title = "Hierarchical Clustering",
    x = "Feature 1",
    y = "Feature 2",
    color = "Cluster"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

p1 + p2
```

## Agreement Between Methods

```{r}
# Calculate agreement (using adjusted Rand index)
library(mclust)

ari <- adjustedRandIndex(km_3_clusters, hclust_3_clusters)
cat("Adjusted Rand Index (K-Means vs Hierarchical):", round(ari, 4), "\n")

# Cross-tabulation
cat("\nCluster Assignment Agreement:\n")
print(table(KMeans = km_3_clusters, Hierarchical = hclust_3_clusters))
```

---

# 7. Practical Guidelines

## Decision Framework for Choosing Optimal k

```{r}
# Create comprehensive decision table
decision_df <- data.frame(
  k = 2:5,
  Elbow_Method = c("No", "Maybe", "YES - Elbow here", "No"),
  Silhouette = c("Moderate (0.60)", "BEST (0.79)", "Good (0.64)", "Decline (0.50)"),
  Davies_Bouldin = c("Higher", "LOWEST", "Low", "Moderate"),
  Calinski_Harabasz = c("Moderate", "Moderate", "MODERATE", "Lower")
)

print(decision_df)

cat("\n=== DECISION SUMMARY ===\n")
cat("Recommended k: 3\n")
cat("Reasoning:\n")
cat("  1. Elbow Method: Clear elbow at k=3\n")
cat("  2. Silhouette Score: Highest at k=3 (0.7883)\n")
cat("  3. Davies-Bouldin: Lowest at k=3 (1.0289)\n")
cat("  4. Calinski-Harabasz: Moderate at k=3 (67.35)\n")
```



---

# 8. Exercises

## Exercise 1: Different Data Distributions

Create your own dataset with 4 clusters and different shapes (e.g., one elongated, one circular). 
How do K-Means and Hierarchical clustering perform?

```{r}
# Template for Exercise 1
exercise_1_data <- data.frame(
  x = c(
    rnorm(30, 0, 0.3),      # Tight circular cluster
    rnorm(30, 5, 0.2),      # Another tight cluster
    runif(30, 8, 12),       # Elongated rectangular cluster
    rnorm(30, 10, 0.5)      # Loose circular cluster
  ),
  y = c(
    rnorm(30, 0, 0.3),
    rnorm(30, 5, 0.2),
    runif(30, 0, 2),
    rnorm(30, 10, 0.5)
  ),
  true_cluster = rep(1:4, each = 30)
)

# Your analysis here...
```

## Exercise 2: Sensitivity to Initialization

Run K-Means 100 times with different random initializations (k=3). 
What is the range of WSS values obtained?

```{r}
# Template for Exercise 2
wss_results <- numeric(100)
for (i in 1:100) {
  km_temp <- kmeans(clusters_3[, 1:2], centers = 3, nstart = 1)
  wss_results[i] <- km_temp$tot.withinss
}

# Your analysis here...
```

## Exercise 3: Optimal k Analysis

Using the evaluation metrics, determine the optimal k for your own dataset.
How do the methods agree or disagree?

```{r}
# Template for Exercise 3
# Generate your own dataset or use existing one

# Step 1: Compute metrics for k=2 to 8
# Step 2: Visualize each metric
# Step 3: Recommend optimal k based on consensus
# Your code here...
```

---

---

# Theoretical Definitions

Below are concise, formal definitions of the key concepts used throughout this lesson. These are intended to provide the theoretical grounding that complements the practical exercises and visualizations.

## Distance Metrics
- **Euclidean distance:** For two points $\mathbf{x},\mathbf{y} \in \mathbb{R}^p$, the Euclidean distance is
  $$d_{E}(\mathbf{x},\mathbf{y}) = \sqrt{\sum_{j=1}^p (x_j - y_j)^2}.$$ 
  It measures straight-line distance in feature space and is sensitive to scale.

- **Manhattan distance (L1):** 
  $$d_{M}(\mathbf{x},\mathbf{y}) = \sum_{j=1}^p |x_j - y_j|.$$
  It measures taxicab distance and is more robust to outliers in some contexts.

- **Cosine distance:** Based on the cosine of the angle between vectors,
  $$d_{cos}(\mathbf{x},\mathbf{y}) = 1 - \frac{\mathbf{x}\cdot\mathbf{y}}{\|\mathbf{x}\|\,\|\mathbf{y}\|}.$$
  It focuses on orientation (angle) rather than magnitude and is useful when feature scale is not informative.

## K-Means Clustering
- **K-Means objective:** Given data $\{\mathbf{x}_i\}_{i=1}^n$ and an integer $k$, K-Means seeks cluster centers $\{\boldsymbol\mu_c\}_{c=1}^k$ that minimize the total within-cluster sum of squares (WSS):
  $$\text{WSS} = \sum_{c=1}^k \sum_{\mathbf{x}_i \in C_c} \|\mathbf{x}_i - \boldsymbol\mu_c\|^2.$$ 
  The standard algorithm alternates between assigning each point to the nearest centroid and updating centroids to the mean of their assigned points.

- **Centroid:** For cluster $C_c$, the centroid is the mean vector
  $$\boldsymbol\mu_c = \frac{1}{|C_c|} \sum_{\mathbf{x}_i \in C_c} \mathbf{x}_i.$$ 

## Within-Cluster Sum of Squares (WSS)
- **Definition:** WSS measures cluster compactness by summing squared Euclidean distances of points to their cluster centroid (see K-Means objective above). Lower WSS indicates tighter clusters but decreases monotonically with larger $k$.

## Silhouette Coefficient
- **Point-level silhouette $s(i)$:** For a point $i$, let $a(i)$ be the average distance to other points in its cluster, and let $b(i)$ be the minimum average distance to points in any other cluster. Then
  $$s(i) = \frac{b(i) - a(i)}{\max\{a(i), b(i)\}},$$
  with $s(i) \in [-1,1]$. Values near $+1$ indicate well-clustered points, values near $0$ indicate points on boundaries, and negative values indicate possible misclassification.

## Dendrogram & Linkage
- **Dendrogram:** A tree representation of hierarchical clustering that records the sequence of merges (agglomerative) or splits (divisive) and the distance (height) at which they occur.

- **Linkage methods:** Rules for computing inter-cluster distances when merging:
  - *Single linkage:* distance between closest pair of points across clusters.
  - *Complete linkage:* distance between farthest pair of points across clusters.
  - *Average linkage:* average pairwise distance between points across clusters.
  - *Ward's method:* chooses merges that minimally increase the total within-cluster variance (equivalently, minimizes increase in WSS).

## External Cluster Validity Indices
- **Davies–Bouldin index (DB):** For each cluster, compute the ratio of within-cluster scatter to between-cluster separation with its most similar cluster; DB is the average of these ratios across clusters. Lower DB indicates better separation.

- **Calinski–Harabasz index (CH):** Ratio of between-cluster variance to within-cluster variance, scaled by degrees of freedom. Higher CH indicates better-defined clusters.

## Convergence & Stopping Criteria (K-Means)
- **Centroid stability:** Stop when $\sum_c \|\boldsymbol\mu_c^{(t)} - \boldsymbol\mu_c^{(t-1)}\|^2 < \varepsilon$.
- **WSS tolerance:** Stop when relative change in WSS between iterations is below a threshold $\delta$.
- **No reassignments:** Stop when cluster assignments do not change between iterations.
- **Max iterations / random restarts:** Practical safeguards include limiting iterations and performing multiple runs with different initializations (selecting result with lowest WSS).

## Elbow Method (Informal)
- Plot WSS versus $k$ and look for a point (elbow) where marginal gains sharply diminish. The elbow is heuristic and should be combined with other metrics and domain knowledge.

---

**Last Updated:** `r Sys.Date()`
