---
title: "R Notebook"
output: html_notebook
---

# Lesson 3: Relationships & Patterns — Interactive Workshop

---

## **Learning Objectives**

By the end of this workshop, you will be able to:
- interpret covariance and correlation coefficients conceptually and mathematically
- istinguish between Pearson's r and Spearman's ρ and know when to use each
- Identify and handle non-linear relationships and outliers
- Detect and diagnose multicollinearity in regression models
- Apply practical workflows to assess model quality and relationship strength

---

## **SECTION 1: COVARIANCE — THE FOUNDATION**

### **1.1 Conceptual Understanding**

**Activity: Quadrant Mapping**

Plot the following 5 data points on a coordinate system. Mark the means X̄ and Ȳ as perpendicular lines. Identify which quadrant each point falls into.

| X | Y |
|---|---|
| 2 | 10 |
| 4 | 20 |
| 6 | 15 |
| 8 | 25 |
| 10 | 30 |

**Questions:**
1. Calculate X̄ and Ȳ
2. Which quadrants contain the most points?
3. Do you expect positive or negative covariance? Why?
4. What does each quadrant's location tell us about the (X−X̄)(Y−Ȳ) product?

**Answer Key:**
- X̄ = 6, Ȳ = 20
- Quadrants I and III dominate → Positive covariance
- Points 1,4,5 in Quadrant I; Point 3 on Y-axis; Point 2 in Quadrant III-ish
- Quadrants I & III contribute positive products; II & IV would contribute negative

---

### **1.2 Calculation Walkthrough**

**Manual Covariance Calculation**

Using the data above, calculate Cov(X,Y) step-by-step:

$$\text{Cov}(X,Y) = \frac{\sum_{i=1}^{n} (X_i - \bar{X})(Y_i - \bar{Y})}{n-1}$$

**Step 1: Calculate deviations**

| X | Y | X−X̄ | Y−Ȳ | (X−X̄)(Y−Ȳ) |
|---|---|------|------|-------------|
| 2 | 10 | -4 | -10 | 40 |
| 4 | 20 | -2 | 0 | 0 |
| 6 | 15 | 0 | -5 | 0 |
| 8 | 25 | 2 | 5 | 10 |
| 10 | 30 | 4 | 10 | 40 |

**Step 2: Sum the products**
$$\sum (X_i - \bar{X})(Y_i - \bar{Y}) = 40 + 0 + 0 + 10 + 40 = 90$$

**Step 3: Divide by (n−1)**
$$\text{Cov}(X,Y) = \frac{90}{4} = 22.5$$

**R Code to Verify:**
```{r}
X <- c(2, 4, 6, 8, 10)
Y <- c(10, 20, 15, 25, 30)
cov(X, Y)  # Should return 22.5
```

---

### **1.3 The Scale Problem**

**Activity: Scale Independence Challenge**

Convert the same data to different units:
- **Dataset A:** X in units, Y in units (original)
- **Dataset B:** X in hundreds, Y in tens

| X_orig | Y_orig | X_scaled | Y_scaled |
|--------|--------|----------|----------|
| 2 | 10 | 0.02 | 1.0 |
| 4 | 20 | 0.04 | 2.0 |
| 6 | 15 | 0.06 | 1.5 |
| 8 | 25 | 0.08 | 2.5 |
| 10 | 30 | 0.10 | 3.0 |

**Task:** Calculate Cov(X_orig, Y_orig) and Cov(X_scaled, Y_scaled)

**R Demonstration:**
```{r}
X_orig <- c(2, 4, 6, 8, 10)
Y_orig <- c(10, 20, 15, 25, 30)

X_scaled <- X_orig / 100
Y_scaled <- Y_orig / 10

cov_orig <- cov(X_orig, Y_orig)
cov_scaled <- cov(X_scaled, Y_scaled)

print(paste("Original Cov:", cov_orig))
print(paste("Scaled Cov:", cov_scaled))
print(paste("Ratio:", cov_orig / cov_scaled))  # Should be 100 * 10 = 1000
```

**Insight:** Same relationship, completely different covariance values!

---

## **SECTION 2: CORRELATION — THE SOLUTION**

### **2.1 Why Correlation?**

**Activity: Unit Cancellation Proof**

Show that correlation is dimensionless:

$$r = \frac{\text{Cov}(X,Y)}{\sigma_X \cdot \sigma_Y} = \frac{\text{(unit X × unit Y)}}{\text{unit X × unit Y}} = \text{dimensionless}$$

Calculate correlation for both original and scaled datasets:

```{r}
# Correlation using built-in function
r_orig <- cor(X_orig, Y_orig)
r_scaled <- cor(X_scaled, Y_scaled)

print(paste("Original r:", r_orig))
print(paste("Scaled r:", r_scaled))
# Both should be identical!
```

**Knowledge Check:**
- Q: Why are the correlation coefficients identical while covariances differ?
- Q: What is the practical advantage of scale invariance?

---

### **2.2 Pearson's r: Range and Proof**

**Activity: Range Boundaries**

**Part A: Perfect Correlation with Self**

For any variable X:
$$r(X,X) = \frac{\text{Cov}(X,X)}{\sigma_X \sigma_X} = \frac{\sigma_X^2}{\sigma_X^2} = 1$$

Test this:
```{r}
X <- c(2, 4, 6, 8, 10)
r_self <- cor(X, X)
print(r_self)  # Should be 1.0
```

**Part B: Perfect Negative Correlation**

For X and −X:
$$r(X,-X) = \frac{\text{Cov}(X,-X)}{\sigma_X \sigma_X} = \frac{-\sigma_X^2}{\sigma_X^2} = -1$$

Test this:
```{r}
X <- c(2, 4, 6, 8, 10)
X_neg <- -X
r_negative <- cor(X, X_neg)
print(r_negative)  # Should be -1.0
```

**Discussion:** Why can r never exceed −1 or +1?

---

### **2.3 Interpreting Pearson's r**

**Activity: Real Dataset Analysis**

Use the provided dataset:

```{r}
# Example: Height vs Weight (synthetic economics student data)
height_cm <- c(165, 172, 158, 180, 175, 168, 182, 170)
income_millions <- c(1.2, 1.8, 0.9, 2.1, 1.9, 1.4, 2.3, 1.6)

r_hw <- cor(height_cm, income_millions)
print(paste("r(height, income) =", round(r_hw, 3)))

# Visualize
plot(height_cm, income_millions, 
     main = paste("r =", round(r_hw, 3)),
     xlab = "Height (cm)", 
     ylab = "Income (millions)")
abline(lm(income_millions ~ height_cm), col = "red")
```

**Interpretation Guide:**
| r Value | Interpretation |
|---------|---|
| ±0.9 to ±1.0 | Very strong linear association |
| ±0.7 to ±0.9 | Strong linear association |
| ±0.5 to ±0.7 | Moderate linear association |
| ±0.3 to ±0.5 | Weak linear association |
| 0 to ±0.3 | Very weak or no linear association |
| 0 | No linear association |

---

## **SECTION 3: NON-LINEARITY — WHEN PEARSON FAILS**

### **3.1 Anscombe's Quartet Experience**

**Activity: Calculate Statistics, Then Plot**

For each of Anscombe's four datasets, calculate:
- mean(X), mean(Y)
- SD(X), SD(Y)
- Pearson's r
- Regression line: Ŷ = a + bX

```{r}
# Anscombe's Quartet
quartet <- data.frame(
  Dataset = rep(c("I", "II", "III", "IV"), each = 11),
  X = c(10, 8, 13, 9, 11, 14, 6, 4, 12, 7, 5,
        10, 8, 13, 9, 11, 14, 6, 4, 12, 7, 5,
        10, 8, 13, 9, 11, 14, 6, 4, 12, 7, 5,
        8, 8, 8, 8, 8, 8, 8, 19, 8, 8, 8),
  Y = c(8.04, 6.95, 7.58, 8.81, 8.33, 9.96, 7.24, 4.26, 10.84, 4.82, 5.68,
        9.14, 8.14, 8.74, 8.77, 9.26, 8.10, 6.13, 3.10, 9.13, 7.26, 4.74,
        7.46, 6.77, 12.74, 7.11, 7.81, 8.84, 6.08, 5.39, 8.15, 6.42, 5.73,
        6.58, 5.76, 7.71, 8.84, 8.47, 7.04, 5.25, 12.50, 5.56, 7.91, 6.89)
)

# Summary statistics by dataset
by(quartet, quartet$Dataset, function(df) {
  cat("\nDataset", df$Dataset[1], "\n")
  cat("Mean X:", mean(df$X), "\n")
  cat("Mean Y:", mean(df$Y), "\n")
  cat("Correlation r:", cor(df$X, df$Y), "\n")
  fit <- lm(Y ~ X, data = df)
  cat("Regression:", round(fit$coefficients[1], 2), "+", round(fit$coefficients[2], 2), "* X\n")
})

# Plot all four
library(ggplot2)
ggplot(quartet, aes(x = X, y = Y)) +
  geom_point(size = 3, color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_wrap(~Dataset, nrow = 2) +
  theme_minimal()
```

**Key Insight:** All four datasets have identical summary statistics, but vastly different relationships!

**Discussion Questions:**
1. Why does Pearson's r miss these differences?
2. What does each dataset teach us about visualization?
3. How would Spearman's ρ perform on Dataset II (curved)?

---

### **3.2 Detecting Non-Linearity**

**Activity: Y = X² Example**

Generate non-linear data and compare Pearson vs. visual assessment:

```{r}
# Generate Y = X^2 + noise
X <- seq(-3, 3, length.out = 30)
Y <- X^2 + rnorm(30, 0, 0.3)

r_pearson <- cor(X, Y)

# Plot and assess
plot(X, Y, main = paste("Y = X² + ε | Pearson r =", round(r_pearson, 3)))
curve(x^2, add = TRUE, col = "green", lty = 2, lwd = 2)
abline(lm(Y ~ X), col = "red", lwd = 2)
legend("top", c("True (X²)", "Linear Fit"), col = c("green", "red"), lty = c(2, 1))
```

**Knowledge Check:**
- Q: Why is Pearson's r close to 0 for Y = X²?
- Q: Would Spearman's ρ be higher? Why?

---

## **SECTION 4: PEARSON vs. SPEARMAN**

### **4.1 Numerical Comparison**

**Activity: Calculate Both Correlations**

For a linear dataset and a non-linear dataset:

```{r}
# Linear data
X_linear <- 1:10
Y_linear <- 2 * X_linear + rnorm(10, 0, 1)

r_pearson_lin <- cor(X_linear, Y_linear, method = "pearson")
r_spearman_lin <- cor(X_linear, Y_linear, method = "spearman")

cat("LINEAR DATA:\n")
cat("Pearson r =", round(r_pearson_lin, 3), "\n")
cat("Spearman ρ =", round(r_spearman_lin, 3), "\n\n")

# Non-linear data (exponential)
X_exp <- seq(0, 4, length.out = 10)
Y_exp <- exp(X_exp) + rnorm(10, 0, 1)

r_pearson_exp <- cor(X_exp, Y_exp, method = "pearson")
r_spearman_exp <- cor(X_exp, Y_exp, method = "spearman")

cat("EXPONENTIAL DATA:\n")
cat("Pearson r =", round(r_pearson_exp, 3), "\n")
cat("Spearman ρ =", round(r_spearman_exp, 3), "\n")
```

**Observation:** Spearman ≈ Pearson for linear; Spearman >> Pearson for nonlinear

---

### **4.2 When to Use Which?**

**Activity: Decision Tree**

For each scenario, decide: Use Pearson r or Spearman ρ?

1. **Scenario A:** Relationship between GDP growth and inflation (both continuous, normally distributed)
   - **Answer:** Pearson r
   - **Why:** Both assumptions met; linear expected; easier interpretation

2. **Scenario B:** Relationship between company rating (1-5 stars) and revenue
   - **Answer:** Spearman ρ
   - **Why:** Rating is ordinal; non-linear possible

3. **Scenario C:** Income vs. years of education (contains outliers)
   - **Answer:** Spearman ρ
   - **Why:** Robust to outliers; captures monotonicity

4. **Scenario D:** Test score vs. study hours (both continuous, linear expected)
   - **Answer:** Pearson r
   - **Why:** Linear relationship; normal distributions likely

---

## **SECTION 5: MULTICOLLINEARITY**

### **5.1 Detecting Multicollinearity**

**Activity: Correlation Matrix Inspection**

Generate correlated predictors and compute correlation matrix:

```{r}
set.seed(42)
n <- 100

# Create correlated variables
X1 <- rnorm(n, mean = 50, sd = 10)
X2 <- 0.85 * X1 + rnorm(n, 0, 5)  # Highly correlated with X1
X3 <- rnorm(n, mean = 100, sd = 15)  # Independent

Y <- 10 + 0.5 * X1 + 0.3 * X2 + 0.1 * X3 + rnorm(n, 0, 2)

# Correlation matrix
cor_matrix <- cor(data.frame(X1, X2, X3, Y))
print(cor_matrix)

# Heatmap
heatmap(cor_matrix)
```

**Interpretation Guide:**
| Correlation | Assessment |
|---|---|
| \|r\| < 0.5 | Low multicollinearity (generally OK) |
| 0.5 ≤ \|r\| < 0.8 | Moderate (investigate) |
| \|r\| ≥ 0.8 | High multicollinearity (problematic) |

---

### **5.2 Variance Inflation Factor (VIF)**

**Activity: Calculate and Interpret VIF**

```{r}
# Fit regression
model <- lm(Y ~ X1 + X2 + X3)
summary(model)

# Calculate VIF
library(car)
vif_values <- vif(model)
print(vif_values)

# VIF Interpretation
cat("\nVIF THRESHOLDS:\n")
cat("VIF < 5: Low multicollinearity (acceptable)\n")
cat("5 ≤ VIF < 10: Moderate (investigate further)\n")
cat("VIF ≥ 10: High multicollinearity (take action)\n")
```

**Action Items Based on VIF:**
- **VIF < 5:** Keep variables as-is
- **5 ≤ VIF < 10:** Consider dimensionality reduction or regularization
- **VIF ≥ 10:** Either drop one variable or combine predictors (PCA)

---

### **5.3 Remedies for Multicollinearity**

**Activity: Compare Model Specifications**

Test approaches:

```{r}
# 1. Original model (high VIF)
model_full <- lm(Y ~ X1 + X2 + X3)

# 2. Drop X2 (highly correlated with X1)
model_reduced <- lm(Y ~ X1 + X3)


# Compare coefficients
cat("Full Model Coefficients:\n")
print(coef(model_full))
cat("\nReduced Model Coefficients:\n")
print(coef(model_reduced))
```

---

## **SUMMARY & TAKEAWAYS**

### **Core Concepts Checklist**

- ☐ Covariance measures joint variability but depends on scale
- ☐ Correlation standardizes covariance to a −1 to +1 range
- ☐ Pearson's r assumes linearity; fails for curved relationships
- ☐ Spearman's ρ captures monotonic trends; robust to non-linearity
- ☐ Always visualize data before computing correlations (Anscombe!)
- ☐ Multicollinearity inflates standard errors and destabilizes coefficients
- ☐ High VIF (>10) signals problematic multicollinearity requiring action

---

## **KNOWLEDGE CHECK — Final Exercise**

**Scenario:** You have economic data on:
- GDP growth (%), continuous, normally distributed
- Unemployment rate (%), continuous, some outliers
- Industry sector (categorical, 1-10 scale)
- Inflation rate (%), continuous

**Questions:**

1. Which pairs would you examine with Pearson's r?
2. Which with Spearman's ρ? Why?
3. If sector code and inflation show r = 0.08, can you conclude they're unrelated? What assumption might be violated?
4. If you fit a model with all four predictors, what multicollinearity diagnostic would you run first?
5. A customer reports r(GDP, Unemployment) varies wildly depending on whether data is in percentages or raw decimals. What's the issue? How do you fix it?

---

## **SOLUTIONS AVAILABLE**

Detailed solutions, R code, and data files available upon request.

---

**End of Workshop**

*Total Time:* 150 minutes  
*Break:* After Section 2 (10 minutes)  
*Exercises:* 12 hands-on activities  
*Code Examples:* 15+ reproducible R snippets
