---
title: "R Notebook"
output: html_notebook
---



```{r}
# Make sure ggplot2 is loaded
library(ggplot2)
# Run this script to generate the 3 CSV files for the workshop
set.seed(12345) # For reproducibility

# ---------------------------------------------------------
# Dataset 1: The Additive Fallacy (Study x Sleep)
# DGP: Rested (D=1) = 60 + 4(Study). All-Nighter (D=0) = 50 + 0(Study)
# ---------------------------------------------------------
n1 <- 500
study_hours <- round(runif(n1, 0, 10), 1)
rest_dummy <- rbinom(n1, 1, 0.5)

# The True Interaction Equation (+ random noise)
exam_score <- 50 + (0 * study_hours) + (10 * rest_dummy) + (4 * study_hours * rest_dummy) + rnorm(n1, 0, 5)
exam_score <- pmin(100, pmax(0, round(exam_score, 1))) # Cap between 0 and 100

df_student <- data.frame(ExamScore = exam_score, StudyHours = study_hours, FullRest = rest_dummy)
write.csv(df_student, "student_performance.csv", row.names = FALSE)

# ---------------------------------------------------------
# Dataset 2: Categorical Interaction (Salary x Degree)
# DGP: Bachelor's (D=0) = 50 + 3(Exp). Master's (D=1) = 60 + 5(Exp)
# ---------------------------------------------------------
n2 <- 500
experience <- round(runif(n2, 0, 15), 1)
degree_dummy <- rbinom(n2, 1, 0.4) # 40% have a Master's

# The True Interaction Equation (+ random noise)
salary <- 50 + (3 * experience) + (10 * degree_dummy) + (2 * experience * degree_dummy) + rnorm(n2, 0, 8)

df_salary <- data.frame(Salary = round(salary, 1), Experience = experience, MastersDegree = degree_dummy)
write.csv(df_salary, "tech_salaries.csv", row.names = FALSE)

# ---------------------------------------------------------
# Dataset 3: Continuous Interaction (Performance = Skill x Motivation)
# ---------------------------------------------------------
n3 <- 500
# Generate raw variables (Uncentered, means around 50)
skill <- rnorm(n3, mean = 50, sd = 10)
motivation <- rnorm(n3, mean = 50, sd = 10)

# True DGP: Complex interaction where motivation acts as a heavy multiplier
performance <- 10 + (0.5 * skill) + (0.5 * motivation) + (0.05 * skill * motivation) + rnorm(n3, 0, 15)

df_productivity <- data.frame(Performance = round(performance, 1), 
                              Skill = round(skill, 1), 
                              Motivation = round(motivation, 1))
write.csv(df_productivity, "employee_productivity.csv", row.names = FALSE)

print("All three datasets generated successfully!")
```
## Activity 1: The Naive Model
load the data and run a standard multiple regression without an interaction term.
look at the StudyHours coefficient and tell you what it means.

```{r}

# 1. Load the data
df_student <- read.csv("student_performance.csv")

# 2. Run the Naive Additive Model
model_naive <- lm(ExamScore ~ StudyHours + FullRest, data = df_student)

# 3. View the results
summary(model_naive)

plot(model_naive)

```
## Activity 2:
Add the interaction term. In R, you can simply replace the + with an *, which automatically includes the main effects and the interaction.


Run the Interaction Model
```{r}
model_interaction <- lm(ExamScore ~ StudyHours * FullRest, data = df_student)
```
View the new results

```{r}
summary(model_interaction)
plot(model_interaction)
```
Compare the R-squared of both models
```{r}
cat("Naive Adjusted R-Squared: ", summary(model_naive)$adj.r.squared, "\n")
cat("Interaction Adjusted R-Squared: ", summary(model_interaction)$adj.r.squared, "\n")
```
## Activity 3: The Residuals
plot the errors (residuals) of both models, color them by the Sleep Condition, and visually prove that the naive model is systematically biased.
```{r}


# 1. Extract the residuals from both models and save them in the dataframe
df_student$resid_naive <- residuals(model_naive)
df_student$resid_interaction <- residuals(model_interaction)

```


Make sure FullRest is treated as a categorical factor for plotting colors
```{r}
df_student$Sleep_Condition <- factor(df_student$FullRest, 
                                     labels = c("All-Nighter (D=0)", "Full Rest (D=1)"))
```
Plot the Residuals of the Naive Additive Model
```{r}
plot_naive_res <- ggplot(df_student, aes(x = StudyHours, y = resid_naive, color = Sleep_Condition)) +
  geom_point(alpha = 0.7, size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5) + # Adds trendlines for the errors
  scale_color_manual(values = c("All-Nighter (D=0)" = "#D55E00", "Full Rest (D=1)" = "#0072B2")) +
  labs(title = "Diagnostic: Naive Model Residuals",
       subtitle = "Systematic Bias: The model makes predictable errors based on sleep",
       x = "Study Hours",
       y = "Residuals (Prediction Error)") +
  theme_minimal(base_size = 14)

print(plot_naive_res)

```

Plot the Residuals of the Interaction Model
```{r}
plot_inter_res <- ggplot(df_student, aes(x = StudyHours, y = resid_interaction, color = Sleep_Condition)) +
  geom_point(alpha = 0.7, size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5) +
  scale_color_manual(values = c("All-Nighter (D=0)" = "#D55E00", "Full Rest (D=1)" = "#0072B2")) +
  labs(title = "Diagnostic: Interaction Model Residuals",
       subtitle = "Random Noise: The interaction term fixes the bias",
       x = "Study Hours",
       y = "Residuals (Prediction Error)") +
  theme_minimal(base_size = 14)

print(plot_inter_res)
```



# Part 2: Dummy Interactions & Geometry
run a categorical interaction, use the coefficients to write the specific regression equations for each group, and plot the non-parallel lines to visualize the intercept and slope shifts.

Dataset Needed: tech_salaries.csv
Variables:

Salary: Base salary in thousands ($k)

Experience: Years of experience (0 to 15)

MastersDegree: 1 (Has Master's) or 0 (Bachelor's Only)


## Activity 1: Extracting the MathInstruct the students to run the interaction model

write the specific $Y = mx + b$ equation for a Bachelor's holder and the equation for a Master's holder based on the output.

```{r}
df_salary <- read.csv("tech_salaries.csv")

model_salary <- lm(Salary ~ Experience * MastersDegree, data = df_salary)

summary(model_salary)

plot(model_salary)
```
## Activity 2: Visualizing the Gap
Now that we have the equations, plot the raw data and fit the regression lines

Convert the dummy variable to a factor for ggplot to color it properly
```{r}

df_salary$Degree_Label <- factor(df_salary$MastersDegree, 
                                 labels = c("Bachelor's (D=0)", "Master's (D=1)"))
```

Plot the data with interaction trendlines
```{r}

plot_salary_interaction <- ggplot(df_salary, aes(x = Experience, y = Salary, color = Degree_Label)) +
  geom_point(alpha = 0.5, size = 2) +
  # geom_smooth with method="lm" automatically fits the interaction lines for us!
  geom_smooth(method = "lm", se = FALSE, linewidth = 2) +
  scale_color_manual(values = c("Bachelor's (D=0)" = "#0072B2", "Master's (D=1)" = "#D55E00")) +
  labs(title = "Tech Salaries by Experience and Degree",
       subtitle = "Visualizing the Intercept Gap and the Slope Difference",
       x = "Years of Experience",
       y = "Salary ($k)",
       color = "Degree Level") +
  theme_minimal(base_size = 14)

print(plot_salary_interaction)
```

# Part 3: Continuous Moderators & The J-N Technique (Instructor Guide)
run a continuous interaction, diagnose multicollinearity using VIF, fix the model using mean-centering, and probe the conditional effects using the Johnson-Neyman technique.

Dataset Needed: employee_productivity.csv
Variables:

Performance: Job performance score (Outcome)

Skill: Technical skill score (0 to 100)

Motivation: Drive/Motivation score (0 to 100)



```{r}
install.packages(c("car", "interactions"))
library(car)
library(interactions)

df_prod <- read.csv("employee_productivity.csv")
```
Run the interaction on the RAW (uncentered) data

```{r}
model_raw <- lm(Performance ~ Skill * Motivation, data = df_prod)
summary(model_raw)


```
Check for Multicollinearity

```{r}
vif(model_raw)
```

Activity 2: Mean-Centering
subtract the mean from both continuous predictors before multiplying them


Mean-Center the continuous predictors
```{r}
# 1. 
df_prod$Skill_c <- df_prod$Skill - mean(df_prod$Skill)
df_prod$Motiv_c <- df_prod$Motivation - mean(df_prod$Motivation)
```


Rerun the model using the centered variables

```{r}
model_centered <- lm(Performance ~ Skill_c * Motiv_c, data = df_prod)
```

Check the VIF again

```{r}
vif(model_centered)
```

```{r}
summary(model_raw)
summary(model_centered)

```
```{r}
plot(model_raw)
```

```{r}
plot(model_centered)
```



## Activity 3: The Johnson-Neyman Plot
Now that the model is centered and stable, we need to report the results. Instead of guessing arbitrary "High/Low" points, we will use the J-N technique to find the exact boundary of significance.


```{r}
# We are testing the effect of Skill (pred) across different levels of Motivation (modx)
jn_results <- sim_slopes(model_centered, pred = Skill_c, modx = Motiv_c, jnplot = TRUE)

# 2. Display the exact Region of Significance boundaries
print(jn_results$jn)

# 3. View the J-N Plot
jn_results$jnplot
```

