---
title: "Workshop 1 — R Refresher: Data, Regression & Reporting"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

---

> **How to use this document:** Follow along with the instructor. Each section
> has empty code chunks — fill them in as we go. At the end of class you will
> have a fully working notebook documenting everything we covered.

---

# Setting Up: Loading Packages and Data

We first load the packages we need and inspect the dataset we will be working
with throughout this workshop.

```{r load-packages}
# Install the wooldridge package if you haven't already:
install.packages("wooldridge")
install.packages("ggplot2")
# Load packages
library(wooldridge)
library(ggplot2)


```

```{r load-data}
# Load the wage1 dataset from the wooldridge package
wage1_dataset = wooldridge::wage1
# Preview the first few rows
head(wage1_dataset)

```

## Exploring the Dataset

```{r explore-data}
# Check the structure of the dataset (variable types, dimensions)
#str(wage1_dataset)

# Get a statistical summary of all variables
summary(wage1_dataset)

```

> **Key variables we care about today:**
>
> - `wage` — hourly wage in USD
> - `educ` — years of education
> - `exper` — years of potential experience
> - `female` — dummy variable (1 = female, 0 = male)

```{r basic-plots}
# Plot the distribution of wages
plot(wage1_dataset$wage, main = "Distribution of Wages", xlab = "Wage", ylab = "Frequency")
# Plot wage against education
plot(wage1_dataset$wage, wage1_dataset$educ,
     main = "Distribution of Wages", xlab = "Wage", ylab = "educ")

```

---

# Running a Simple Linear Regression

We start with the simplest possible model: regressing wage on education only.

```{r simple-lm}
# Fit a simple linear regression: wage ~ educ
fited_model = lm(wage ~ educ, data = wage1_dataset)
# Display the summary output
summary_small = summary(fited_model)


```

## Reading the Output — Line by Line

Use the space below to annotate what each part of the output means.
Fill this in as the instructor walks through it.

| Output element       | What it is                          | Link to theory    |
|----------------------|-------------------------------------|-------------------|
| `Estimate` (Intercept) |                                   |                   |
| `Estimate` (educ)    |                                     | $\hat{\beta}_1$   |
| `Std. Error`         |                                     |                   |
| `t value`            |                                     |                   |
| `Pr(>|t|)`           |                                     |                   |
| `R-squared`          |                                     |                   |
| `F-statistic`        |                                     |                   |

---

# Running a Multiple Linear Regression

Now we add more regressors to control for other factors.

```{r multiple-lm}
# Fit a multiple regression: wage ~ educ + exper + female
mlr_fit = lm(wage1_dataset$wage ~ wage1_dataset$educ + wage1_dataset$exper + wage1_dataset$female + wage1_dataset$married + wage1_dataset$exper + wage1_dataset$tenure, data = wage1_dataset) 

summary_mlr_fit <- summary(mlr_fit)
summary(mlr_fit) -> summary_mlr_fit
# Display the summary output
plot(mlr_fit)

```

## Comparing the Two Models

```{r compare-models}
# Extract and compare R-squared values between the two models
summary_mlr_fit$r.squared
summary_small$r.squared
# get th eadjusted R-squared
summary_mlr_fit$adj.r.squared
summary_small$adj.r.squared

```

> **Question to think about:** The coefficient on `educ` changed when we added
> `exper` and `female`. Why? What does this tell us about the simple model?
> *(Hint: think about Omitted Variable Bias from Lesson 1.)*

---

# Extracting and Using Model Components

```{r model-components}
# Extract the coefficient vector
vector_of_coefficients <- coef(mlr_fit)
vector_of_coefficients

# Extract residuals
residuals <- resid(mlr_fit)

# Extract fitted (predicted) values
fitted_values <- fitted(mlr_fit)

# Verify R-squared manually from residuals
rss <- sum(residuals^2)  # Residual Sum of Squares
tss <- sum((wage1_dataset$wage - mean(wage1_dataset$
wage))^2)  # Total Sum of Squares

r_squared_manual <- 1 - (rss / tss)

#adjusted R-squared
n = nrow(wage1_dataset)  # Number of observations
k = length(coef(mlr_fit)) - 1  # Number of predictors

adjusted_r_squared_manual <- 1 - ((1 - r_squared_manual) * (n - 1) / (n - k - 1))

adjusted_r_squared_manual
summary_mlr_fit$adj.r.squared

```

---

# Your Turn — Independent Exercise

## Task

Fit a new regression model of your choice using variables from `wage1`.
You must include **at least three regressors**.

```{r exercise-model}
# Your regression here


```

## Interpret Your Results

Write your interpretation below (use plain text, not code):

**Coefficient of interest:**

**Interpretation:**

**Is the coefficient statistically significant? At what level?**

**What might be omitted from this model? In which direction would the bias go?**


---

*Workshop 1 · Applied Statistics · `r format(Sys.Date(), "%B %Y")`*
