---
title: "Workshop 1 — R Refresher: Data, Regression & Reporting"
subtitle: "Complete Reference Version"
author: "Applied Statistics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 4.5
)
```

------------------------------------------------------------------------

> **About this document:** This is the complete reference version of
> Workshop 1. It contains all code, outputs, and interpretations. Keep
> it — it is your blueprint for running and reporting regressions in R
> throughout the course.

------------------------------------------------------------------------

# Setting Up: Loading Packages and Data

```{r load-packages}
# install.packages("wooldridge")  # run once if not installed
library(wooldridge)
library(ggplot2)     # for nicer plots (optional but recommended)
```

```{r load-data}
data("wage1")        # load dataset from the wooldridge package
head(wage1)          # first 6 rows
```

## Exploring the Dataset

```{r explore-data}
str(wage1)           # variable types and dimensions
summary(wage1)       # min, max, mean, quartiles for every variable
```

**Key variables:**

| Variable | Description                        |
|----------|------------------------------------|
| `wage`   | Hourly wage in USD                 |
| `educ`   | Years of education                 |
| `exper`  | Years of potential work experience |
| `female` | 1 = female, 0 = male               |
| `tenure` | Years with current employer        |

```{r basic-plots}
# Distribution of wages
hist(wage1$wage,
     main = "Distribution of Hourly Wages",
     xlab = "Wage (USD/hour)",
     col  = "steelblue",
     border = "white")

# Wage vs Education
plot(wage1$educ, wage1$wage,
     main = "Wage vs Years of Education",
     xlab = "Years of Education",
     ylab = "Hourly Wage (USD)",
     pch  = 16,
     col  = adjustcolor("steelblue", alpha.f = 0.5))
abline(lm(wage ~ educ, data = wage1), col = "red", lwd = 2)
```

> **What to notice:** Wages are right-skewed — a few high earners pull
> the mean up. Education has a positive association with wages, but
> there is a lot of scatter. That scatter is what $\varepsilon$
> captures.

------------------------------------------------------------------------

# Simple Linear Regression

## Fitting and Summarising the Model

```{r simple-lm}
model_simple <- lm(wage ~ educ, data = wage1)
summary(model_simple)
```

## Reading the Output — Complete Annotation

### Coefficients table

```         
            Estimate Std. Error t value Pr(>|t|)    
(Intercept) -0.90485    0.68497  -1.321    0.187    
 educ        0.54136    0.05325  10.167   <2e-16 ***
```

| Element | Value | Meaning |
|----|----|----|
| `Estimate` (Intercept) | −0.905 | Predicted wage when `educ = 0`. Not economically meaningful here. |
| `Estimate` (educ) | **0.541** | $\hat{\beta}_1$: one extra year of education is associated with \$0.54/hr more in wages, *ceteris paribus* (holding nothing else constant here — simple regression). |
| `Std. Error` (educ) | 0.053 | Estimated standard deviation of $\hat{\beta}_1$ across repeated samples. |
| `t value` | 10.15 | $t = \hat{\beta}_1 / SE = 0.541 / 0.053$. Distance from zero in standard errors. |
| `Pr(>|t|)` | \< 2e-16 | p-value. Probability of seeing $|t| \geq 10.15$ if $H_0: \beta_1 = 0$ were true. Essentially zero. |

### Model-level statistics

```         
Residual standard error: 3.378 on 524 degrees of freedom
Multiple R-squared:  0.1648,   Adjusted R-squared:  0.1632
F-statistic: 103.1 on 1 and 524 DF,  p-value: < 2.2e-16
```

| Element | Value | Meaning |
|----|----|----|
| `Residual std. error` | 3.378 | Estimate of $\sigma = \sqrt{\text{Var}(\varepsilon)}$. Average prediction error in wage units (USD/hr). |
| `R-squared` | 0.165 | Education alone explains 16.5% of the variance in wages. Most variation is unexplained — it's in $\varepsilon$. |
| `F-statistic` | 103.1 | Joint test: is the model as a whole useful? Equivalent to t-test here (one regressor). |

> **Recall from Lesson 1:** $R^2 = 0.165$ is not "low" or "high" in
> isolation. The important question is: what is in the remaining 83.5%?
> Ability? Experience? Industry? That is the OVB question.

------------------------------------------------------------------------

# Multiple Linear Regression

## Fitting the Model

```{r multiple-lm}
model_multi <- lm(wage ~ educ + exper + female, data = wage1)
summary(model_multi)
```

## Interpretation

```         
Coefficients:
             Estimate Std. Error t value Pr(>|t|)
(Intercept)  -2.1238     0.7279  -2.917  0.00367 **
educ          0.5705     0.0499  11.434  < 2e-16 ***
exper         0.0250     0.0115   2.167  0.03060 *
female       -1.8116     0.2649  -6.838  1.7e-11 ***
```

**Coefficient interpretations:**

-   **`educ` = 0.571:** Holding experience and gender fixed, one
    additional year of education is associated with \$0.57/hr higher
    wages. *(Increased slightly from simple model — why? Experience and
    gender are not perfectly correlated with education.)*

-   **`exper` = 0.025:** Holding education and gender fixed, one
    additional year of experience is associated with \$0.025/hr higher
    wages. Statistically significant but economically modest.

-   **`female` = −1.812:** Holding education and experience fixed,
    female workers earn on average \$1.81/hr less than male workers.
    Highly significant. This is a raw gender wage gap conditional on our
    controls — not a causal estimate of discrimination.

## Comparing Models

```{r compare-models}
cat("Simple model  R²:", round(summary(model_simple)$r.squared, 4), "\n")
cat("Multiple model R²:", round(summary(model_multi)$r.squared, 4), "\n")
cat("Multiple model Adj. R²:", round(summary(model_multi)$adj.r.squared, 4), "\n")
```

> **Why did R² jump?** Adding `exper` and `female` explained additional
> variance. The adjusted R² accounts for the extra parameters — if it
> also increases, the new variables are genuinely useful.

------------------------------------------------------------------------

# Extracting Model Components

```{r model-components}
# Coefficient vector
coef(model_multi)

# First 10 residuals
head(residuals(model_multi), 10)

# First 10 fitted values
head(fitted(model_multi), 10)

# Manual verification of R²
SS_res <- sum(residuals(model_multi)^2)
SS_tot <- sum((wage1$wage - mean(wage1$wage))^2)
R2_manual <- 1 - SS_res / SS_tot
cat("Manually computed R²:", round(R2_manual, 4), "\n")
```

## Residual Diagnostics (Preview)

```{r residual-plot}
plot(fitted(model_multi), residuals(model_multi),
     main = "Residuals vs Fitted Values",
     xlab = "Fitted Values",
     ylab = "Residuals",
     pch  = 16,
     col  = adjustcolor("steelblue", alpha.f = 0.4))
abline(h = 0, col = "red", lwd = 2)
```

> **What to look for:** Under GM5 (homoskedasticity), the spread of
> residuals should be roughly constant across fitted values. Does it
> look constant here? We will revisit this in a later class.

------------------------------------------------------------------------

# Confidence Intervals

```{r confidence-intervals}
# 95% confidence intervals for all coefficients
confint(model_multi, level = 0.95)
```

> **Recall from Lesson 1:** The 95% CI for `educ` tells us the range of
> $\hat{\beta}_{educ}$ values consistent with the data. If it excludes
> zero, we reject $H_0: \beta_{educ} = 0$ at the 5% level — consistent
> with the t-test result.

------------------------------------------------------------------------

# Key R Functions — Quick Reference

| Function                     | Purpose                         |
|------------------------------|---------------------------------|
| `data("wage1")`              | Load built-in dataset           |
| `head(df)`                   | First 6 rows                    |
| `str(df)`                    | Structure: types and dimensions |
| `summary(df)`                | Descriptive statistics          |
| `lm(y ~ x1 + x2, data = df)` | Fit OLS regression              |
| `summary(model)`             | Full regression output          |
| `coef(model)`                | Extract coefficients            |
| `residuals(model)`           | Extract residuals               |
| `fitted(model)`              | Extract fitted values           |
| `confint(model)`             | Confidence intervals            |
| `plot(fitted, resid)`        | Residual diagnostic plot        |

------------------------------------------------------------------------

# R Markdown Quick Reference

| Syntax                      | Output                     |
|-----------------------------|----------------------------|
| `**bold**`                  | **bold**                   |
| `*italic*`                  | *italic*                   |
| `` `code` ``                | `code`                     |
| `# Heading 1`               | Large heading              |
| `## Heading 2`              | Sub-heading                |
| ```` ```{r chunk-name} ```` | Start a code chunk         |
| `echo=FALSE`                | Run chunk, hide code       |
| `include=FALSE`             | Run chunk, hide everything |
| `eval=FALSE`                | Show code, don't run       |
